prolfqua - An R-package for Proteomics Differential Expression Analysis.

### 2022-05-05 inst/MyArticle/paper.Rmd
Improve description of UML diagrams.

---
title: prolfqua - An R-package for Proteomics Differential Expression Analysis.
author: 
  - Paolo Nanni$^{\star}$
  - Jonas Grossmann$^{\star}$
  - Maria d'Errico$^{\star}$
  - Ralph Schlapbach$^{\star}$
  - Christian Panse$^{\star}$
  - Witold E. Wolski $^{\star}$^[[Functional Genomics Center Zurich (FGCZ)](https://fgcz.ch) - Swiss Federal Institute of Technology in Zurich \| University of Zurich; Winterthurerstrasse 190, CH-8057 Zurich, Switzerland. Correspondence$^\star$\texttt{wew@fgcz.ethz.ch}]
output: 
  bookdown::pdf_document2:
    toc: true
bibliography: bibliography.bib
editor_options: 
  chunk_output_type: console
abstract: |
    Protein mass spectrometry is widely used for quantitative proteomics studies, relative protein quantification, and differential expression analysis of proteins. Nevertheless, there is still a need for a flexible and easy-to-use application programmer interface in R that transparently supports a variety of well principled statistical procedures. The package can model simple experimental designs with a single explanatory variable and complex experiments with multiple factors and hypothesis testing.
    The `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package integrates, using an R6 application programmer interface, essential steps of the mass spectrometry-based differential expression analysis workflow: quality control, data normalization, protein aggregation, statistical modeling, hypothesis testing, and sample size estimation. 
     Furthermore, the package implements benchmark functionality that can help to compare data acquisition, data preprocessing, or data modeling methods using a gold standard dataset. Finally, we show that the implemented methods allow sensitive and specific differential expression analysis. 
     The R package is available through https://github.com/fgcz/prolfqua.

@@

(ref:LFQData) Class Diagram for LFQData and related classes. The LFQData_Plotter class uses the LFQData class to implement methods for plotting. Similarily the LFQData_Stats, LFQData_Summary, reference the LFQData class and group methods for variance and sample size estimation or for summarizing peptide and protein counts.

@@

__R__ linear model and linear mixed effect models allow modeling parallel designs, repeated measurements, factorial designs, and many more. __R__'s formula interface for linear models is flexible, widely used, and well documented ([@faraway2016extending], page : ). This makes it easy to reproduce an analysis performed with `r BiocStyle::Githubpkg("fgcz/prolfqua")` in other statistical programming languages. We implemented features specific to high throughput experiments, such as the experimental Bayes variance and $p$-value moderation, which utilizes the parallel structure of the protein measurements and the analysis [@Ritchie2015]. We also compute probabilities of differential protein regulation based on peptide level models [@Suomi2017bEnhanced]. We use R6 to model the statistical modeling functionality in `r BiocStyle::Biocpkg("prolfqua")` (see Figure \@ref(fig:ContrastUML)).

(ref:ContrastUML) UML diagram of modeling and contrast related classes. Different strategies, e.g., _lm_, _lmer_, and _glm_ can be used to fit various types of models using a model builder (see code). All classes computing contrasts implement the _Contrast_ interface.

>>>>

---
title: prolfqua - An R-package for Proteomics Differential Expression Analysis.
author: 
  - Witold E. Wolski $^{\star}$^[[Functional Genomics Center Zurich (FGCZ)](https://fgcz.ch) - Swiss Federal Institute of Technology in Zurich \| University of Zurich; Winterthurerstrasse 190, CH-8057 Zurich, Switzerland. Correspondence$^\star$\texttt{wew@fgcz.ethz.ch}]
  - Paolo Nanni$^{\star}$
  - Jonas Grossmann$^{\star}$
  - Maria d'Errico$^{\star}$
  - Ralph Schlapbach$^{\star}$
  - Christian Panse$^{\star}$
output: 
  bookdown::pdf_document2:
    toc: true
bibliography: bibliography.bib
editor_options: 
  chunk_output_type: console
abstract: |
    Protein mass spectrometry is widely used for quantitative proteomics studies, relative protein quantification, and differential expression analysis of proteins. Nevertheless, there is still a need for a flexible and easy-to-use application programmer interface in R that transparently supports a variety of well principled statistical procedures. The package can model simple experimental designs with a single explanatory variable and complex experiments with multiple factors and hypothesis testing.
    The `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package integrates, using an R6 application programmer interface, essential steps of the mass spectrometry-based differential expression analysis workflow: quality control, data normalization, protein aggregation, statistical modeling, hypothesis testing, and sample size estimation. 
     Furthermore, the package implements benchmark functionality that can help to compare data acquisition, data preprocessing, or data modeling methods using a gold standard dataset. Finally, we show that the implemented methods allow sensitive and specific differential expression analysis. 
     The R package is available through https://github.com/fgcz/prolfqua.

@@

(ref:LFQData) Class Diagram of classes representing the proteomics data. The LFQData class encapsulates the data. Other classes are decorating the LFQData class with additional functionality. The LFQData_Plotter class uses the LFQData class to implement methods for data visualization. Similarly, the LFQData_Stats, LFQData_Summary, reference the LFQData class, group methods for variance and sample size estimation, or summarizing peptide and protein counts.

@@

__R__ linear model and linear mixed effect models allow modeling parallel designs, repeated measurements, factorial designs, and many more. __R__'s formula interface for linear models is flexible, widely used, and well documented [@faraway2016extending]. We use the formula interface to specify the models which should make it easy to reproduce an analysis performed with `r BiocStyle::Githubpkg("fgcz/prolfqua")` in other statistical programming languages. In addition we implemented features specific to high throughput experiments, such as the experimental Bayes variance and $p$-value moderation, which utilizes the parallel structure of the protein measurements and the analysis [@Ritchie2015]. We also compute probabilities of differential protein regulation based on peptide level models [@Suomi2017bEnhanced]. We used R6 classes encapsulate the statistical modelling functionality in `r BiocStyle::Biocpkg("prolfqua")` (see Figure \@ref(fig:ContrastUML)). We did specify a contrast interface (ContrastInterface), and there are various implementations allowing to perform differential expression analysis given linear or mixed effect models (Contrast), variance shrinkage (ContrastModerated), or impute contrasts in case of observations missing for an entire group of samples (ContrastSimpleImpute) can also be used to encapsulate and integrate differential expression analysis results of external tools such as SAINTExpress or proDA.

(ref:ContrastUML) UML diagram of modeling and contrast related classes. Different strategies, e.g., _lm_, _lmer_, and _glm_ can be used to fit various types of models. The model builder method fit's the statistical model given the data and a strategy. The obtained model can then be used for analysis of variance or to estimate contrasts. All classes estimating the contrasts implement the _Contrast_ interface. The variance of the contrasts can be moderated using the `ContrastModerated` class.
